A significance test in forward stepwise regression

### 2014-03-07 paper/paper.tex
Possible fix of categorical variable problem
Fixed some issues with glinternet, prevent groups already counted from being added again

% "Title of the paper"
\title{A significance test in forward stepwise model selection}

@@

% indicate corresponding author with \corref{}
\author{\fnms{Jonathan} \snm{Taylor}\corref{Jonathan E. Taylor}\ead[label=e1]{jonathan.taylor@stanford.edu}\thanksref{t1}
\and \fnms{Joshua} \snm{Loftus}
\and ???}
\thankstext{t1}{Supported in part by NSF grant DMS 1208857 and
AFOSR grant 113039.}

@@

\address{Department of Statistics\\  Stanford University\\ Sequoia
Hall \\390 Serra Mall\\ Stanford, CA 94305, U.S.A.\\ \printead{e1}}

@@

\begin{abstract}
% perhaps de-emphasize group lasso and emphasize forward stepwise?
  We apply the methods of \cite{tests:adaptive} and
  \cite{significance:lasso} on significance tests for penalized
  regression to forward stepwise model selection. For the $k$th
  variable to enter the model, previous work relied on an asymptotic
  null distribution that breaks down when grouped variables are
  entering the model and is difficult to derive. We iteratively apply
  an exact null distribution for the first (potentially grouped)
  variable on the residual from previous steps. The resulting method
  has the strengths of stepwise selection, for example parallel
  computation, but also remedies the problem of inflated test
  statistics and over-fitting.
\end{abstract}

%%%
%%% - Gotta change these
%%%
\begin{keyword}[class=AMS]
\kwd[Primary ]{62M40}
%\kwd{}
\kwd[; secondary ]{62H35}
\end{keyword}

\begin{keyword}
\kwd{penalized regression}
\kwd{convex analysis}
\kwd{least squares}
\kwd{Gaussian processes}
%\kwd{extreme value theory}
\end{keyword}

@@

Forward stepwise regression is a stochastic model selection procedure
that begins with an empty model and adds the best predictor variable
in each step.  Classical significance tests fail when a model has been
selected this way and tend to be anti-conservative.  Recently,
\cite{significance:lasso} found a novel test statistic with an
appropriate null distribution that behaves well when a model has been
selected using the lasso \citep{tibshirani:lasso}.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other adaptive regression
problems.  The present work explores the behavior of those test
statistics for models selected by forward stepwise procedures and
works out some of the details involved in applying these methods to
models with grouped variables.  Our test statistic can be
used for valid significance testing when computed from the same
data as the model selection.  The resulting method can
be more statistically efficient than validation on held-out data, and
also more computationally efficient than penalized methods with
regularization parameters chosen by cross-validation.

(\todo Change this paragraph as the sections are completed)
In Section \ref{sec:stepwise} we establish notation and describe our
forward stepwise procedure. Section \ref{sec:testing} briefly reviews
the parts of \cite{significance:lasso} and \cite{tests:adaptive}
relevant to our significance test, and describes the group lasso which
we require for applying the test with grouped
variables. Simulation results in Section \ref{sec:simulations} using
various stopping rules for the forward stepwise procedure--including
some from \cite{sequential:fdr}--appear promising.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special case.

Label the outcome variable of $n$ i.i.d. measurements $y \in \real^n$. Let an integer $G \geq 1$ 
be the number of groups. For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables in group $g$. Define $p = \sum_{g=1}^G
p_g$ the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%Throughout most of this paper we assume orthogonality within groups, that is
%$X_g^T X_g = I_{p_g \times p_g}$ for all $g$. ( \todo Take this out if
%we don't actually need it).  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

We allow for the possibility $p > n$ in which case \ref{eq:gmodel} is
generically underdetermined. In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries \citep{donoho:pursuit}. In the rest of this paper we refer to
variable groups $X_g$ as noise variables if $\beta_g$ is a zero vector
and as true predictor variables if $\beta_g$ has nonzero entries.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure, given in Algorithm
\ref{algo:fs}. Note that to enable our p-value computations,
\textit{max.steps} should be at most $\min (n, G) - 1$. In our
implementation we treat the active set $A$ as an ordered set to easily
track the order of variables entering the model.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$step \leftarrow 1$ \KwTo $max.steps$}{
  $g^* \gets \argmax_{g \in A^c} \norm{X_g^T r_{g-1}}_2 / w_g$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  \For{$g \in A^c$}{
    $X_g \gets (I - X_{g^*}X_{g^*}^\dagger) X_g$;
  }
  $r_g \gets $\lsfitResidual{$r_{g-1}, X_{g^*}$}\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

Subset selection refers to the problem of choosing a subset of the $p$
potential variables to include in a model. Since there are $2^p$ such
choices, exhaustive search is computationally infeasible when $p$ is
large. If exhaustive search were possible, performing a large
search runs the risk of over-fitting unless model complexity is
appropriately penalized. Algorithm \ref{algo:fs} produces a much
smaller set of potential models, at most \textit{max.steps}. However,
it is a greedy algorithm so the set of models it produces may not
contain the best possible model. This is an inherent shortcoming which
we do not address in this paper.

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some classical approaches can be posed as optimization
criteria which stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
where we have written $\{ \beta_s : s = 1, \ldots, max.steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). In the current paper we provide
accurate p-values when adding the first noise variable group in the forward
stepwise sequence, and it is natural to consider using these p-values
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section \ref{sec:simulations}.

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a parallel literature to the
compressed sensing literature \textit{(cite some here)} with theorems
stating that forward stepwise can exactly select the true model under
some stringent conditions involving quantities like the sparsity of
the true model and the coherence of the design matrix. We do not
pursue this any further because we find empirically that forward
stepwise can work well in much less restrictive situations than those
guaranteed by theorems.

To explore the performance of forward stepwise empirically we
conducted simulations with results shown in ~\ref{fig:fwdstepsims}

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
%\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
%\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\label{fig:fwdstepsim}
\end{center}
\end{figure}

@@

\todo Update this section to match/complement \citep{tests:adaptive}

In the ordinary least squares setting, a significance test for a
single variable can be conducted by comparing the drop in residual
sums of squares (RSS) to a $\chi^2_1$ distribution. Similarly, when
adding a group of $k$ variables we can compare the drop in RSS to a
$\chi^2_k$ random variable. This generally does not work when the
variable to be added has been chosen by a method that uses the data,
and in particular it fails for forward stepwise procedures adding
the ``best'' (e.g. most highly correlated) predictor in each step. In
that case, the test statistic (drop in RSS) does not match the
theoretical null distribution even when the null hypothesis is
true. \cite{significance:lasso} introduced a new test statistic based
on the knots in the lasso solution path. They derived a simple
asymptotic null distribution, proved a type of convergence under broad
``minimum growth'' conditions, and demonstrated in simulations that
the test statistic closely matches its asymptotic distribution even in
finite samples.  That work marked an important advance in the problem
of combining inference with model selection.  The current paper
extends some of their results to the group lasso \citep{grouplasso}.
In the process we had to derive an exact finite sample null
distribution for the test statistic, this appears in
\cite{tests:adaptive}.

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we will
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso. (\todo Change this if I can find $\lambda_k$).

The \emph{group lasso estimator} is the following solution to a convex
problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt p_g$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.

@@

\todo Expand, include simulation results (copy latex for
includegraphics etc), do the simulation comparing stopping rules

\begin{itemize}
  \item Show null and non-null p-value-by-step plots for several examples
  \item Discuss one of these carefully (perhaps with a noise variable
    entering before a signal variable), for several fixed stopping points
  \item Compare a few stopping rules, e.g. naive one(s),
    TailStop/HybridStop, AIC/BIC/Cp
  \item (the stopping comparisons are not done yet, but I am guessing
    they will be comparable, perhaps AIC/BIC/Cp will be better, and in
    that case look for other advantages of ours to mention,
    e.g. interpretability for being based on p-values rather than a
    complexity penalty)

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9_corr0p1.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\section{Real data example}
\todo Simulation with non-Gaussian design matrix (from AIDS data probably)

>>>>

% "Title of the paper"
\title{A significance test in forward stepwise regression}

@@

% indicate corresponding author with \corref{}
%\author{\fnms{Jonathan} \snm{Taylor}\corref{Jonathan E. Taylor}\ead[label=e1]{jonathan.taylor@stanford.edu}\thanksref{t1}
%\and \fnms{Joshua} \snm{Loftus}}
%\thankstext{t1}{Supported in part by NSF grant DMS 1208857 and AFOSR grant 113039.}

@@

\address{Department of Statistics\\  Stanford University\\ Sequoia
Hall \\390 Serra Mall\\ Stanford, CA 94305, U.S.A.\\ }
%\printead{e1}}

@@

\begin{abstract}
% perhaps de-emphasize group lasso and emphasize forward stepwise?
  We extend the methods developed by \cite{significance:lasso} and
  \cite{tests:adaptive} on significance tests for penalized
  regression to stepwise model selection by iteratively applying
  the global null hypothesis test on the residual in a forward stepwise
  procedure. While not fully theoretically justified, the resulting method
  has the computational strengths of stepwise selection and partially
  solves the problem of inflated test statistics due to model selection.
  We illustrate the flexibility of this method by applying it to
  novel specialized applications of forward stepwise to a hierarchically
  constrained interactions model and generalized additive models.
\end{abstract}

%%%
%%% - Gotta change these??????????
%%%
\begin{keyword}[class=AMS]
\kwd[Primary ]{62M40}
%\kwd{}
\kwd[; secondary ]{62H35}
\end{keyword}

\begin{keyword}
\kwd{forward stepwise}
\kwd{model selection}
\kwd{significance test}
\end{keyword}

@@

Forward stepwise regression is a classical model selection procedure
that begins with an empty model and sequentially adds the best predictor
variable in each step. Classical significance tests fail
when a model
has been selected this way and tend to be anti-conservative.
In the lasso setting, \cite{significance:lasso} derived a novel test
statistic along with a correct asymptotic distribution, making possible
valid inferences after model selection.
\cite{tests:adaptive} modified and extended those results to the
group lasso \citep{grouplasso} and other penalized regression
problems, but only under the global null hypothesis.
One of the strengths of these test statistics is that they can be
used for valid significance testing when computed on the same
data used for model selection, eliminating the need for data splitting.
The present work iteratively applies the global null test
of \cite{tests:adaptive} for each step in forward stepwise selection,
and works out some of the details necessary for
models with grouped variables. The resulting method can
be more statistically efficient than validation on held-out data, and
more computationally efficient than penalized methods with
regularization parameters chosen by cross-validation.

In Section~\ref{sec:stepwise} we establish notation and describe the
forward stepwise algorithm used in most of the paper.
Section~\ref{sec:testing} briefly reviews
the parts of \cite{significance:lasso} and \cite{tests:adaptive}
relevant to our significance test, and describes the group lasso which
we require for applying the test with grouped variables.
Simulation results in Section~\ref{sec:simulations} show
empirically that our method performs well in settings where forward
stepwise itself performs well, and that
various stopping rules using our test statistic--including
some from \cite{sequential:fdr}--appear promising. In
Section~\ref{sec:applications} we apply
the method to several variants of forward
stepwise tailored to models with interactions and generalized additive
models, as well as to a real data example.
(\todo Change this paragraph if necessary as the sections are completed)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Background and notation}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{../figs/pubmed.pdf}
\caption{\small \it Forward stepwise enjoys widespread use among
practitioners}
\label{fig:pubmed}
\end{center}
\end{figure}

As a classical method dating back about half a century
(see \cite{classical:selection} for a review),
forward stepwise regression has not received much attention in
recent years in the theoretical statistics community. But it continues
to be widely used by practitioners.
For example, search results on PubMed for forward stepwise
(summarized in Figure~\ref{fig:pubmed}) show that many recent
papers mention the method and there is an increasing trend over time.
Its popularity among researchers continues despite the fact that it
invalidates inferences using the standard $\chi^2$ or $F$-tests.

Some classical attempts to address this issue include Monte Carlo
estimation of tables of adjusted $F$-statistic values \citep{mc:ftoenter}
and permutation statistics \citep{permutation:stop}. Aside from the works
this paper is based on, there have been other recent attempts to do
inference after model selection. Most of these make use of subsampling
\citep{meinshausen:buhlmann} or data splitting \citep{wasserman:roeder}.
Our approach allows use of the full data and does not require the
extra computation involved in subsampling.
Before describing the full approach we first introduce notation and
specify our implementation of forward stepwise, which is slightly
different from the most commonly used versions.
(\todo Check R's \textit{step} for the group criterion)

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway.
To emphasize this we will use $g, h$ as indices rather than the usual
$i, j$ throughout. Since single variables can be considered groups of
size 1, our general exposition includes non-grouped situations as a special
case. Our variant of forward stepwise uses a different method than
usual for choosing which group to add because we are using the
same objective in the forward stepwise procedure that was used to
derive the test statistics in penalized regression settings.

Denote $y \in \real^n$ for $n$ i.i.d. measurements of the outcome variable.
Let an integer $G \geq 2$ be the number of groups of explanatory variables.
For each $1 \leq g \leq G$ the design matrix encoding the
$g$th group is the $n \times p_g$ matrix denoted $X_g$, where $p_g$ is
the number of individual variables or columns in group $g$. When a group
encodes a categorical variable as binary indicators for the levels of
that variable, we use the full encoding with a column for every level.
Although this introduces collinearity, our method does not require the
design matrix to have full rank.

Define $p = \sum_{g=1}^Gp_g$,
the total number of individual variables, so $p = G$ in the
case where all groups have size 1. Let $X$ be the matrix constructed
by column-binding the $X_g$, that is  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation*}
X = \begin{pmatrix} X_1 & X_2 & \cdots & X_G  \end{pmatrix}
\end{equation*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We assume throughout that the individual columns of $X$ are scaled to
have unit 2-norm (we do not scale groups, this is accomplished later
by weighting).
With each group we associate the $p_g \times 1$ coefficient vector
$\beta_g$, and write $\beta$ for the $p \times 1$ vector constructed
by stacking all of the $\beta_g$ in order.  Finally, our model for the
response is
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gmodel}
y & = X \beta + \sigma \epsilon \\
   & = \sum_{g=1}^G X_g \beta_g + \sigma \epsilon
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\epsilon$ is noise. Unless otherwise specified we assume
i.i.d. Gaussian noise $\epsilon \sim N(0, I_{n \times n})$ and that
$\sigma$ is known.

We allow for the possibility $p > n$ in which case \eqref{eq:gmodel} is
generically underdetermined. In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries. In the rest of this paper we refer to variable groups $X_g$
as noise groups if $\beta_g$ is a zero vector and as true or signal
groups if $\beta_g$ has any nonzero entries. We refer to the number
of such nonzero groups as the \textit{sparsity} of the model, and
denote this $k := \# \{ g : \beta_g \neq 0 \}$.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will mostly use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure.

\subsection{Description of the algorithm}

First, the user must specify the maximum number of steps allowed,
which we denote \textit{steps} (consistent with the \textit{R} function
for stepwise). To enable our test statistic computations,
\textit{steps} should be at most $\min (n, G) - 1$, but it is
computationally desirable to set it as low as possible while still being
larger than the (unknown) sparsity of $\beta$.
This way forward stepwise has a chance of recovering all the nonzero
coefficients of $\beta$ and then terminating without performing much
additional computation.
In our implementation we treat the active set $A$ as an ordered list to easily track the order of groups entering the model.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ groups of variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{\small \it Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$s \gets 1$ \KwTo $steps$}{
  $g^* \gets \argmax_{g \in A^c} \{ \norm{X_g^T r_{s-1}}_2 / w_g \}$\;
  $P \gets I - X_{g^*}X_{g^*}^\dagger$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  \For{$g \in A^c$}{
    $X_g \gets P X_g$\;
  }
  $r_s \gets P r_{s-1}$\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

In the common case where all groups are single variables this version
coincides with the standard implementations of forward stepwise.
We will now use forward stepwise to refer specifically to
Algorithm~\ref{algo:fs} unless otherwise specified.

\subsection{Performance of forward stepwise}

Among model selection procedures, forward stepwise is one which performs
variable selection: from a potentially large set of variables it chooses
a subset to include in the model. The most general form of variable
selection is called subset selection, a procedure which searches among
all $2^G$ subsets of $G$ variable groups and picks the best model.
This exhaustive search is computationally infeasible when $G$ is
large, and when possible it still
runs the risk of over-fitting unless model complexity is
appropriately penalized (as in \eqref{eq:subsetregress} below).
Forward stepwise produces a much
smaller set of potential models, with cardinality at most
\textit{steps}. However
it is a greedy algorithm, so the set of models it produces may not
contain the best possible model. This is an inherent shortcoming of
forward stepwise procedures and should be kept in mind when choosing
between model selection methods.

So far we have left open the question of choosing among the models in
the forward stepwise sequence, i.e. when to stop stepping
forward. Some approaches for this problem can be posed as optimization
criteria which stop at the step minimizing
\begin{equation}
\begin{aligned}
\label{eq:subsetregress}
\frac{1}{2} \| y - X \beta_s \|_2^2 + \lambda \pen(\beta_s)
\end{aligned}
\end{equation}
(\todo describe construction of $\beta$ after finding $A$)
where we have written $\{ \beta_s : s = 1, \ldots, steps \}$ as
the sequence of models outputted by forward stepwise. The function
$\pen(\beta)$ is a penalty on model complexity usually taken to be the
number of nonzero entries of $\beta$. Proposals for $\lambda$ include
2 ($C_p$ of \cite{CP}, AIC of \cite{AIC}), $\log(n)$ (BIC of \cite{BIC}), and
$2\log(p)$ (RIC of \cite{RIC}). Stopping rules based on classical test
statistics have also been used, so it is natural to consider using
the new test statistics of \cite{significance:lasso} or
\cite{tests:adaptive}
to choose a model. \cite{sequential:fdr} examined some stopping rules
using the asymptotic p-values of \cite{significance:lasso} and showed
their stopping rules control false discovery rate--the expected
proportion of noise variables among variables declared significant
\citep{fdr}. We explore this further in Section~\ref{sec:simulations}.

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a small segment of the
compressed sensing literature \citep{donoho:pursuit, cai:wang:omp}
with results stating that forward stepwise (often referred to in that
literature as Orthogonal Matching Pursuit or OMP) can exactly select
the true model
under some stringent conditions involving quantities like the sparsity
of the true model and the coherence of the design matrix. The
coherence $\mu(X)$ of a matrix $X$ with columns scaled to have unit
2-norm is defined as

\begin{equation}
  \mu := \mu(X) = \max_{i \neq j} \{ | \innerp{ X_i }{ X_j } | \}
\end{equation}

Recalling that $k$ is the sparsity of $\beta$, typical results in this
literature say that if $k < (1/\mu + 1)/2$ and the nonzero
coefficients of $\beta$ are sufficiently large then forward stepwise
recovers $\beta$ with high probability. The coherence condition is
necessary to guarantee exact recovery \citep{cai:wang:xu:sharp} in the
sense that it is possible to construct counterexamples with
$k = (1/\mu + 1)/2$. We refer the reader to the literature for details.
For our purposes the
conditions required to guarantee exact recovery are usually too
stringent. Simulations show empirically that forward
stepwise can work well even when it is not working perfectly, and that
it does so under a wide range of conditions.

\begin{figure}
\begin{center}
\subfigure[Ternary design]{
\label{fig:fsternary}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/ternary_p2n_n100_lower1p4_upper1p4.pdf}}
\hspace{-15pt}
\subfigure[Gaussian design with correlation]{
\label{fig:fsgausscorr}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows the results of the simulation
  using design matrices with i.i.d. ternary entries taking values $0,
  \pm 1$ with equal probability. The left panel shows the results
  with design matrices of standard Gaussian entries with
  equi-correlation of 0.2. }
\label{fig:fwdstepsim}
\end{center}
\end{figure}

For various sparsity levels $k$, Algorithm~\ref{algo:fs} was run on
various data sets with \textit{steps} equal to $k$ and the ``average
power'' was calculated as the number of true variables in the active
set after $k$ steps divided by $k$. Nonzero coefficients had magnitude
in a range of multiples of $\gamma := \sqrt{2\log(G)}$, e.g. in $[1.4
\gamma, 1.6 \gamma]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. After computing the
coherence of these design matrices, some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\gamma$. These simulations are far
from the stringent conditions required by theory to guarantee perfect
recovery, but the performance, while not perfect, may still be good
enough for some purposes.

@@

\todo Reorganize and update this section \\

In the ordinary least squares setting, a significance test for a
single variable can be conducted by comparing the drop in residual
sums of squares (RSS) to a $\chi^2_1$ distribution. Similarly, when
adding a group of $k$ variables we can compare the drop in RSS to a
$\chi^2_k$ random variable. This generally does not work when the
group to be added has been chosen by a method that uses the data,
and in particular it fails for forward stepwise procedures adding
the ``best'' (e.g. most highly correlated) predictor in each step. In
that case, the test statistic (drop in RSS) does not match the
theoretical null distribution even when the null hypothesis is
true. \cite{significance:lasso} introduced a new test statistic based
on the knots in the lasso solution path. They derived a simple
asymptotic null distribution, proved a type of convergence under broad
``minimum growth'' conditions, and demonstrated in simulations that
the test statistic closely matches its asymptotic distribution even in
finite samples.  That work marked an important advance in the problem
of combining inference with model selection.  \cite{tests:adaptive}
extended that work to the group lasso \citep{grouplasso} and other
problems, and modified the test statistic to one with an exact finite
sample distribution under the global null hypothesis.

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we will
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso.

The \emph{group lasso estimator} is a solution to the following convex
problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt {p_g}$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.

@@

\todo Clean up and reorganize \\
\todo Write a little more exposition \\
\todo Compare to AIC and other classical stopping rules? \\

\begin{figure}[h]
\begin{center}
\subfigure[Gaussian design]{
\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/gaussian_size1_n100_p200_g200_k10_lower1p8_upper1p9_corr0p1.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\end{center}
\end{figure}

@@

\section{Applications}
\label{sec:applications}
We now turn to applying forward stepwise and examining the behavior of
our test statistic in several unique settings, including a real data
example.
